Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-04 Manuscript/manuscript.tex
Corrected section Residence times and Conclusions.

Equilibration of \ce{Ca^{2+}} binding to POPC bilayer takes hundreds of
nanoseconds in MD simulations with CHARMM36 and Slipids force fields~\cite{javanainen17}, suggesting that
simulations with the length of several microseconds would be required
to study the binding phenomena in equilibrium. However, the density profiles of \ce{Ca^{2+}}
from the ECC-lipids simulation in Fig.~\ref{fig:cacl-dens} show a relatively shallow
minimum at the interface, suggesting a smaller barrier for the exchange of calcium ions.
To quantify the ion binding exchange in lipid bilayer simulations, we calculated 
the residence times of ions within the distance of 0.3~nm from any oxygen atom
belonging to a lipid molecule in a bilayer.

The residence time histograms of \ce{Ca^{2+}}
in POPC bilayer are shown in Fig.~S\ref{fig:hist_residence_times} in SI 
from the ECC-lipids simulation with CaCl$_2$ concentration of 1~M
and from the CHARMM36 simulation with 450~mM concentration of
ECC-ions taken from the previous study \cite{javanainen17}
(the simulation data was downloaded from Ref.~\citenum{zenodo.259376}).
\todo{JOE: Plot a histogram of residence times based on our simulations and simulation from \citenum{javanainen17} and put it into SI.
}
\todo{It might be also better to do the analysis from ECC-lipids simulation with concentration closer
to the data in literature, i.e. 450mM.}
%Based on Fig. S17 in that work, Ref.~\citenum{javanainen17},
%we took the simulation with the apparent highest fluctuations of the number of bound calcium,
%which can be obtained at \url{https://doi.org/10.5281/zenodo.259376}
%and anaylsed the residence times.
% simulation: (Matti C36 with CaClMK, 800ns total):
Approximately 20\% of the \ce{Ca^{2+}} residence times in the CHARMM36 simulation are longer than
the lenght of the trajectory ($800\,\mathrm{ns}$),
and only less than 60\% of the bound residence times are shorter than the half
of the simulation length~(400~ns).
Even longer residence times are observed
in the other simulations with CHARMM36 and Slipids models reported in the previous work~\cite{javanainen17}.
At least order of magnitude faster exchange is observed in the
ECC-lipids model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are 
shorter than $60\,\mathrm{ns}$, % exactly $53\,\mathrm{ns}$
and the longest observed residence time in the ECC-lipids simulation is $141\,\mathrm{ns}$, 
which is well below the total length of the simulation (200~ns).
The exchange of \ce{Na+} ions to a POPC membrane is
an order of magnitude faster, giving 90\%
of the residence times smaller than~$1\,\mathrm{ns}$,
the longest residence time being~$6\,\mathrm{ns}$. 
All the results are in line with the experimental estimation that 
the residence time of \ce{Ca^{2+}} at the individual PC headgroup 
is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}.

@@

In conclusion, the results from ECC-lipids model suggest that the
exchange of bound calcium occurs within $\sim$100ns timeframe,
which is much faster than previously reported~\cite{javanainen17}.
The sodium exhibit even faster exchange.
This suggests that simulations with the length of few hundred
nanoseconds are sufficient to simulate equilibrium ion binding
to PC lipid bilayers, when realistic force fields are used.
This has not been the case with the available lipid force fields,
which overestimate the binding streght of the cations \cite{javanainen17,catte16}.
%Such a finding changes the point of view on binding of calcium to PC membranes from
%a very strong long-term stable binding with rare exchanges to 
%a relatively frequent exchange of clacium cations 
%in equilibrium between membrane and solvent.

@@

By using the electrometer concept we show that the binding of \ce{Na^+} and \ce{Ca^2+} ions
to a POPC lipid bilayer can be accurately described with the classical MD simulation 
force field, where the electronic polarization is implicitly included using 
the electronic continuum correction (ECC)~\cite{leontyev11}.
The proposed ECC-lipids model is a significant improvement over 
other available lipid models, which all overestimate specific cation binding affinities~\cite{catte16}.  
While the structural details of a POPC lipid bilayer simulated with the ECC-lipids
model agree with experiments with the comparable accuracy to the other state of the art lipid models,
it also reproduces the experimental lipid head group order parameter responses to
the cationic surfactant, NaCl and CaCl$_2$ concentrations.

The good agreement with experiments enables the atomic resolution 
interpretation of NMR experiments by using MD simulations.
In agreement with previous interpretations of experimental data \cite{hauser76,hauser78,herbette84,binder02},
the Ca$^{2+}$ ions mainly interact with phosphate oxygens.
However, the stoichiometry of the binding is significantly more complicated 
than in the previous interpretation of the NMR data based on 
the ternary complex model, where one calcium binds to two POPC molecules~\cite{altenbach84}.
The complexes of one calcium bound to two lipids are the most probable also in the
ECC-lipids model, but the complexes of one or three lipids per one calcium
were observed to be almost equally likely. While the success of the ternary
compex model is understandable based on the simulation results,
a simple binding model cannot detec the complex binding
oberved in the simulation.

The improved cation binding behaviour to POPC bilayer pave the way for
simulations of complex biochemical systems with correctly described
electrostatic interactions in the vicinity of cellular membranes.
The ECC-lipids model is build by scaling the partial charges
and the LJ-radius of the headgroup, glycerol backbone and carbonyl atoms of
the Lipid14 POPC model~\cite{dickson14} with the constant factor
of 0.8 and 0.89, respectively. While the Lipid14 model
is compatible with the AMBER force field family, the compatibility of
the ECC-lipids model may be compromised due to the changes in intermolecular
interactions of the scaled atoms. On the other hand, a fully consistent
ECC-force field should include the correction also in other than lipid
molecules, including water. The work toward this direction and the extension
to other lipid molecules and force fields is left for the future studies.

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
P.J. acknowledges support from the Czech Science Foundation (grant no. 16-01074S) 
and 600 from the Academy of Finland via the FiDiPro award.
\todo{Add acknowledgments of Jiri Kolafa.}
\end{acknowledgments}

@@

>>>>

Equilibration of \ce{Ca^{2+}} binding to POPC bilayer takes hundreds of
nanoseconds in MD simulations with current state of the art force fields,
e.g. CHARMM36 and Slipids force fields~\cite{javanainen17}, suggesting that
simulations at a temporal scale of several microseconds are required
for studies of cation binding in equilibrium. 
This is in line with the density profiles of \ce{Ca^{2+}} 
from simulations in Fig.~\ref{fig:cacl-dens}, 
which all show a region of a very small calcium density 
around the level of cholines in POPC (i.e. around $2.5\,\mathrm{nm}$)
with the exception of ECC-lipids, 
which exhibits only a relatively shallow local
minimum at that level.
This suggests that such a smaller barrier for the exchange of calcium ions 
shall lead to their faster exchange with the solvent.
To quantify the exchange of ions between the membrane and solvent in simulations, we calculated 
residence times of ions bound to the membrane.
We again use the definition for ions binding to the membrane as in the above section~\ref{sec:affinity},
i.e. ion is bound if it is within 0.3~nm from any oxygen atom
belonging to a lipid molecule in a bilayer.

The histograms of residence times of \ce{Ca^{2+}} in a POPC bilayer 
at $450\,\mathrm{mM}$ concentration of \ce{CaCl_2} in water before adsorption to phospholipids
from simulations with ECC-lipids 
and CHARMM36 (taken from previous study \cite{javanainen17}, 
simulation data available at Ref.~\citenum{zenodo.259376})
are shown in Fig.~S\ref{fig:hist_residence_times} in SI. 
%Based on Fig. S17 in that work, Ref.~\citenum{javanainen17},
%we took the simulation with the apparent highest fluctuations of the number of bound calcium,
%which can be obtained at \url{https://doi.org/10.5281/zenodo.259376}
%and anaylsed the residence times.
In the CHARMM36 simulation, approximately 20\% of the \ce{Ca^{2+}} residence times are longer than
the length of the trajectory ($800\,\mathrm{ns}$).
With such a simulation, which is hence relatively short for the employed model,
we can merely estimate an upper bound that 
only less than 60\% of the bound residence time 
is formed by cations interacting with the phospholipid membrane 
for less then half of the simulation length~(400~ns).
Even longer residence times are observed
in the other simulations with CHARMM36 and Slipids models reported in the previous work~\cite{javanainen17}.

On the other hand, at least an order of magnitude faster exchange is observed in the
ECC-lipids model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are 
shorter than $60\,\mathrm{ns}$, % exactly $53\,\mathrm{ns}$
and the longest observed residence time in the ECC-lipids simulation is $141\,\mathrm{ns}$, 
which is well below the total length of the simulation used for analysis, 200~ns.
The exchange of \ce{Na+} ions to a POPC membrane is
another order of magnitude faster, giving 90\%
of the residence times smaller than~$1\,\mathrm{ns}$,
the longest residence time being~$6\,\mathrm{ns}$. 
Note that the results from both analyzed models are in line with the experimental estimation that 
the residence time of \ce{Ca^{2+}} at the individual PC headgroup 
is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}.

@@

In conclusion, the results from the ECC-lipids model suggest that the
exchange of calcium between membrane and solvent occurs within $\sim 100\,$ns timeframe,
which is much faster than previously reported~\cite{javanainen17}.
Sodium cations exhibit even a faster exchange.
This suggests that simulations with the length of a few hundreds of
nanoseconds are sufficient to simulate ion binding
to phospholipid bilayers in equilibrium, when realistic force fields are used.
This has not been the case with the available lipid force fields,
which overestimate the binding strength of the cations \cite{javanainen17,catte16}.
%Such a finding changes the point of view on binding of calcium to PC membranes from
%a very strong long-term stable binding with rare exchanges to 
%a relatively frequent exchange of clacium cations 
%in equilibrium between membrane and solvent.

@@

We used the electrometer concept to show that the binding of \ce{Na^+} and \ce{Ca^2+} ions
to a POPC lipid bilayer can be accurately described with a classical MD simulation 
force field, where the electronic polarization is implicitly included using 
the electronic continuum correction (ECC)~\cite{leontyev11}.
The proposed ECC-lipids model is a significant improvement over 
other available lipid models, which all overestimate specific cation binding affinities~\cite{catte16}.  
While the structural details of a POPC lipid bilayer simulated with the ECC-lipids
model agree with experiments with an accuracy comparable to the other state of the art lipid models,
it also reproduces the experimental lipid head group order parameter responses to
a cationic surfactant, NaCl and CaCl$_2$ concentrations.

The good agreement with experiments enables us to 
interpret NMR experiments in atomistic details using MD simulations.
In line with previous interpretations of experimental data \cite{hauser76,hauser78,herbette84,binder02},
Ca$^{2+}$ ions interact mainly with phosphate oxygens.
However, the stoichiometry of calcium binding is significantly more complicated 
than in the previous interpretation of the NMR data based on 
the ternary complex model, where one calcium binds to two POPC molecules~\cite{altenbach84}.
The complexes of one calcium bound to two lipids are the most probable also in the
ECC-lipids model, but the complexes of one or three lipids per one calcium
were observed to be almost equally likely. While the success of the ternary
compex model is understandable based on the simulation results,
a simple binding model cannot detect 
the complex nature of calcium binding to phospholipid bilayers
oberved in the simulation.

The accurate description of cation binding to POPC bilayer paves the way for
simulations of complex biochemical systems with realistically described
electrostatic interactions in the vicinity of cellular membranes.
The ECC-lipids model is build by scaling the partial charges
and the Van der Waals radii of the headgroup, glycerol backbone and carbonyl atoms of
the Lipid14 POPC model~\cite{dickson14} with the constant factor
of 0.8 and 0.89, respectively. While the Lipid14 model
is compatible with the AMBER force field family, the compatibility of
the ECC-lipids model may be compromised due to the changes in intermolecular
interactions of the scaled atoms. On the other hand, a fully consistent
ECC-force field should include the correction also in other than lipid
molecules, including water. The work toward this direction and the extension
to other lipid molecules and force fields is left for future studies.

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
P.J. acknowledges support from the Czech Science Foundation (grant no. 16-01074S) 
and 600 from the Academy of Finland via the FiDiPro award.
\todo{Add acknowledgments of Jiri Kolafa.}
\todo{Joe: Thank Metacentrum.}
\end{acknowledgments}

@@

\begin{figure}[tbp!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/histogram_bound_times_ECC-lipids_346mM_CaCl.pdf} \\
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/histogram_bound_times_Charmm36_450mM_CaCl_Matti.pdf}
  \caption{\label{fig:hist_residence_times}
   Histograms of residence times of \ce{Ca^{2+}} in a POPC bilayer 
   at $450\,\mathrm{mM}$ concentration of \ce{CaCl_2} in water before adsorption to phospholipids
   from simulations with ECC-lipids/ECC-ions (top)
   and CHARMM36/ECC-ions (bottom), which was taken from previous study \cite{javanainen17} 
   (simulation data available at Ref.~\citenum{zenodo.259376}).
   The maxima on the x-axis represent the lengths of the simulations used for analysis. 
   For the ECC-lipids model, 90\% of the residence times of \ce{Ca^{2+}} in a POPC membrane are 
   shorter than $60\,\mathrm{ns}$, % exactly $53\,\mathrm{ns}$
   with the longest observed residence time being $141\,\mathrm{ns}$, 
   which is well below the total length of the simulation (200~ns).
   This is, however not true for the simulation with CHARMM36 model,
   where there are several calcium cations with their residence time 
   apparently limited by the length of the simulation. 
   With such a simulation, which is relatively short for the employed model,
   we can merely estimate an upper bound that 
   only less than 60\% of the bound residence time 
   is formed by cations interacting with the phospholipid membrane 
   for less then half of the simulation length~(400~ns).
  }
\end{figure}
